---
output:
  md_document:
    variant: markdown_github
    toc: true
    toc_depth: 3
    includes:
      before_body: inst/header.md
  html_document:
    toc: true
    toc_depth: 3
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(comment = "#>")
options(crayon.enabled = TRUE)
options(crayon.colors = 256)
crayon::num_colors(forget = TRUE)
```

```{r, comment="", results="asis", echo = FALSE}
old.hooks <- fansi::set_knit_hooks(knitr::knit_hooks)
```

## Features

* Calls an R function, with arguments, in a subprocess.
* Copies function arguments to the subprocess and copies the return value
  of the function back, seamlessly.
* Copies error objects back from the subprocess, including a stack trace.
* Shows and/or collects the standard output and standard error of the
  subprocess.
* Supports both one-off and persistent R subprocesses.
* Calls the fucntion synchronously or asynchronously (in the background).
* Can call `R CMD` commands, synchronously or asynchronously.
* Can call R scripts, synchronously or asynchronously.
* Provides extensible `r_process`, `rcmd_process` and `rscript_process`
  R6 classes, based on `processx::process`.

## Installation

Install the stable version from CRAN:

```r
install.packages("callr")
```

## Synchronous, one-off R processes

Use `r()` to run an R function in a new R process. The results are
passed back seamlessly:

```{r}
library(callr)
r(function() var(iris[, 1:4]))
```

### Passing arguments

You can pass arguments to the function by setting `args` to the list of
arguments. This is often necessary as these arguments are explicitly
copied to the child process, whereas the evaluated function cannot
refer to variables in the parent. For example, the following does
not work:

```{r, error = TRUE}
mycars <- cars
r(function() summary(mycars))
```

But this does:

```{r}
r(function(x) summary(x), args = list(mycars))
```

Note that the arguments will be serialized and saved to a file,
so if they are large R objects, it might take a long time for the
child process to start up.

### Using packages

You can use any R package in the child process, just make sure to
refer to it explicitly with the `::` operator. For example, the following
code creates an [igraph](https://github.com/igraph/rigraph) graph
in the child, and calculates some metrics of it.

```{r}
r(function() { g <- igraph::sample_gnp(1000, 4/1000); igraph::diameter(g) })
```

### Error handling

callr` copies errors from the child process back to the main R session:

```{r include = FALSE}
options(rlib_error_always_trace = TRUE)
```

```{r error = TRUE}
r(function() 1 + "A")
```

callr sets the `.Last.error` variable, and after an error you can inspect
this for more details about the error, including stack traces both from the
main R process and the subprocess.

```{r}
.Last.error
```

The error objects has two parts. The first belongs to the main process,
and the second belongs to the subprocess.

`.Last.error` also includes a stack trace, that includes both the main
R process and the subprocess:

```{r include = FALSE}
# Remove most of the artificial knitr frames from the trace
local({
  tr <- .Last.error.trace
  rcall <- which(vapply(
    tr$calls,
    function(x) length(x) >= 1 && identical(x[[1]], quote(r)),
    logical(1)))
  tr$ignore <- c(tr$ignore, list(c(1L, rcall - 1L)))
  print(rcall)
  env <- as.environment("org:r-lib")
  assign(".Last.error.trace", tr, envir = env)
})
```

```{r}
.Last.error.trace
```




```{r include = FALSE}
options(rlib_error_always_trace = FALSE)
```

The top part of the trace contains the frames in the main process, and the
bottom part contains the frames in the subprocess. Frame 38. the
anonymous function that is copied to the subprocess.

### Standard output and error

By default, the standard output and error of the child is lost,
but you can request callr to redirect them to files, and then
inspect the files in the parent:

```{r}
x <- r(function() { print("hello world!"); message("hello again!") },
  stdout = "/tmp/out", stderr = "/tmp/err"
)
readLines("/tmp/out")
readLines("/tmp/err")
```

With the `stdout` option, the standard output is collected and can
be examined once the child process finished. The `show = TRUE` options
will also show the output of the child, as it is printed, on the console
of the parent.

## Background R processes

`r_bg()` is similar to `r()` but it starts the R process in the background.
It returns an `r_process` R6 object, that provides a rich API:

```{r}
rp <- r_bg(function() Sys.sleep(.2))
rp
```

This is a list of all `r_process` methods:

```{r}
ls(rp)
```

These include all methods of the `processx::process` superclass and the
new `get_result()` method, to retrieve the R object returned by the
function call. Some of the handiest methods are:

* `get_exit_status()` to query the exit status of a finished process.
* `get_result()` to collect the return value of the R function call.
* `intertupt()` to send an interrupt to the process. This is equivalent
  to a CTRL+C keypress, and the R process might ignore it.
* `is_alive()` to check if the process is alive.
* `kill()` to terminate the process.
* `poll_io()` to wait for any standard output, standard error, or the
  completion of the process, with a timeout.
* `read_*()` to read the standard output or error.
* `suspend()` and `resume()` to stop and continue a process.
* `wait()` to wait for the completion of the process, with a timeout.

## Multiple background R processes and `poll()`

Multiple background R processes are best managed with the
`processx::poll()` function that waits for events (standard output/error or
termination) from multiple processes. It returns as soon as one process
has generated an event, or if its timeout has expired. The timeout is in
milliseconds.

```{r}
rp1 <- r_bg(function() { Sys.sleep(1/2); "1 done" })
rp2 <- r_bg(function() { Sys.sleep(1/1000); "2 done" })
processx::poll(list(rp1, rp2), 299)
rp2$get_result()
processx::poll(list(rp1), 1000)
rp1$get_result()
```

## Persistent R sessions

`r_session` is another `processx::process` subclass that represents a
persistent background R session:

```{r}
rs <- r_session$new()
rs
```

`r_session$run()` is a synchronous call, that works similarly to `r()`,
but uses the persistent session. `r_session$call()` starts the function
call and returns immediately. The `r_session$poll_process()` method or
`processx::poll()` can then be used to wait for the completion or other
events from one or more R sessions, R processes or other
`processx::process` objects.

Once an R session is done with an asynchronous computation, its
`poll_process()` method returns `"ready"` and the `r_session$read()`
method can read out the result.

```{r}
rs$run(function() runif(10))
rs$call(function() rnorm(10))
rs
rs$poll_process(2000)
rs$read()
```

## Running `R CMD` commands

The `rcmd()` function calls an `R CMD` command. For example, you can
call `R CMD INSTALL`, `R CMD check` or `R CMD config` this way:

```{r}
rcmd("config", "CC")

#>$stdout
#>[1] "clang\n"
#>
#>$stderr
#>[1] ""
#>
#>$status
#>[1] 0
```

This returns a list with three components: the standard output, the standard
error, and the exit (status) code of the `R CMD` command.

## License

MIT © Mango Solutions, RStudio
