---
title: Persistent External R Sessions
output:
  github_document:
    toc: true
    toc_depth: 3
always_allow_html: yes
editor_options:
  markdown:
    wrap: sentence
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  cache = FALSE,
  out.width = "100%"
)
asciicast::init_knitr_engine(
  echo = TRUE,
  echo_input = FALSE,
  startup = quote(library(cli))
)
options(
  asciicast_theme = "pkgdown",
  asciicast_cols = 80
)
```

```{asciicast include = FALSE}
options(width = 80)
```

## Introduction

`callr::r_session` is a class for a persistent R session that runs in the
background and you can send commands to it.
It extends the `processx::process` class, so all methods of that class are
still available for use.

## Starting and terminating an R session

Use `r_session$new()` to start an R session.
By default `r_session$new()` blocks, it does not return until the R
session is up and running and ready to run R commands.
If an error happens during process startup, including an R error, then
`r_session$new()` throws an error.
A blocking `r_session$new()` waits at most `wait_timeout` milliseconds for R
to start up.
`wait_timeout` is by default 3000 milliseconds, which should be plenty.
Typically R starts up in about 100-300 milliseconds.

```{asciicast start}
library(callr)
system.time(rs <- r_session$new())
```

```{asciicast start-1}
rs
```

```{asciicast start-2}
rs$get_state()
```

To terminate an R session, call its `$close()` method:

```{asciicast close}
rs$close()
rs
```

Just like `processx::process` objects, `r_session` objects have a finalizer,
and they will be terminated when the R object that represents them is garbage
collected.

### Non-blocking startup

If you don't want to wait for the session to start up, then use
`wait = FALSE` in `r_session$new()`.
If you do that, `r_session$new()` will still error if the OS cannot start
up the R process, but R errors will be reported asynchronously.

```{asciicast start-bg}
system.time(rs2 <- r_session$new(wait = FALSE))
```

```{asciicast start-bg-1}
rs2
```

```{asciicast start-bg-2}
rs2$get_state()
```

You can use `processx::poll()` to wait for the R session being ready, with a
timeout.
The timeout can also be 0ms for a quick check without waiting.
This lets you do extra work in the main process while the R process is
starting up.
It also lets you start up multiple processes concurrently, see the next
section.

```{asciicast start-bg-3}
processx::poll(list(rs2), 3000)
```

The important part of the output is the `process` connection. This will be
`"ready"` if the R process is up and running, or if an error happened.
It will be `"timeout"` if it is not yet ready.

`output` and `error` will be `"ready"` if the R process emitted something to
its standard output and standard error, respectively.
Usually these will be `"silent"` because we suppress R output during startup
with command line options.
This can be changed via the `options` argument and `r_session_options()`.

Once `processx::poll()` reports a `"ready"` `process` connection, you can call the
`r_session$read()` method to see if the startup was successful.

If `r_session$$read()` reports "201 STARTED", it is ready to run R code:

```{asciicast start-bg-4}
rs2$read()
```

## Options

You can use the `options` argument of `r_session$new()` to change the default
startup options.
`options` must be a named list and it is best to create its value with
`r_session_options()`.
Pass the options you want to change as named arguments to
`r_session_options()`.
See `?r_session_options` for the details.

Here is an example that uses the `load_hook` option to run extra code
right after R has started up:

```{asciicast options}
opts <- r_session_options(
  load_hook = quote({ message("I am running!"); Sys.sleep(1) })
)
rs3 <- r_session$new(wait = FALSE, options = opts)
processx::poll(list(rs3), 3000)
```

```{asciicast options-2}
rs3$read_error()
```

```{asciicast options-3}
rs3$poll_process(3000)
```

```{asciicast options-4}
rs3$read()
```

Use the `$poll_process()` method to poll only the process for being ready,
without polling the standard output and error.
Note, however, that if the process generates enough output on stdout
or stderr that fills the pipe buffer between the processes, then it will
stop running, until the main process reads the pipe.

```{asciicast options-5, include = FALSE}
rs3$close()
```

## Running multiple R sessions

If you need to start several R sessions quickly, then it is best to use
`wait = FALSE` and then `processx::poll()` for all processes until they are
all ready.

```{r include = FALSE}
num_procs <- 4
```

```{asciicast multiple}
num_procs <- 4
procs <- tibble::tibble(
  session = replicate(num_procs, r_session$new(wait = FALSE), simplify = FALSE),
  started_at = Sys.time(),
  start_result = list(NULL)
)
limit <- Sys.time() + as.difftime(5, units = "secs")
while ((now <- Sys.time()) < limit &&
       any(vapply(procs$session, function(p) p$get_state(), "") == "starting")) {
  timeout <- as.double(limit - now, units = "secs")
  pr <- processx::poll(procs$session, as.integer(timeout * 1000))
  lapply(seq_along(pr), function(i) {
    if (pr[[i]][["process"]] == "ready") {
      procs$start_result[[i]] <<- procs$session[[i]]$read()
    }
  })
}
Sys.time() - procs$started_at
```

```{asciicast multiple-2}
procs$start_result
```

## Running code

`r_session` objects have three methods to run R code:

* `$run()` is synchronous and omits standard output and error.
* `$run_with_output()` is synchronous and collects standard output and error.
* `$call()` is asynchronous and collects standard output and error.

Let's use the `r num_procs` R sessions created above to demonstrate them.

`$run()` is the simplest:

```{asciicast run}
procs$session[[1]]$run(function() glue::glue("I am process {Sys.getpid()}."))
```

`$run_with_output()` has the output as well:

```{asciicast run-with-output}
procs$session[[1]]$run_with_output(function() {
  message("I am process ", Sys.getpid(), ".")
  head(mtcars)
})
```

`$call()` starts running the function, but does not wait for the result:

```{asciicast call}
invisible(lapply(procs$session, function(p) {
    p$call(function() {
      Sys.sleep(runif(1) * 2)
      glue::glue("I am process {Sys.getpid()}.")
    })
}))
procs$session
```

Use `processx::poll()` to wait for one or more sessions to finish their
job:

```{asciicast call-1}
pr <- processx::poll(procs$session, 5000)
pr
```

Then you can use the `$read()` method to read out the result
(or error, if a failure happened):

```{asciicast call-2}
for (i in seq_along(pr)) {
  if (pr[[i]][["process"]] == "ready") {
    cat("Process ", i, " is ready:\n")
    print(procs$session[[i]]$read())
  }
}
```

To wait for all processes to be ready, you can use a loop that is similar
to the one we used above to start them.
You might find this helper function useful as a starting point:

```{asciicast call-3}
wait_for_sessions <- function(sess, timeout = 5000) {
  result <- vector("list", length(sess))
  is_busy <- function() {
    vapply(sess, function(s) s$get_state() == "busy", logical(1))
  }
  limit <- Sys.time() + as.difftime(timeout / 1000, units = "secs")
  while ((now < Sys.time()) < limit && any(busy <- is_busy())) {
    towait <- as.integer(as.double(limit - now, units = "secs") * 1000)
    pr <- processx::poll(sess[busy], towait)
    for (i in seq_along(pr)) {
      if (pr[[i]][["process"]] == "ready") {
        result[busy][[i]] <- sess[busy][[i]]$read()
      }
    }
  }
  result
}
wait_for_sessions(procs$session)
```


Errors from a `$run()` are turned into errors in the main process:

```{asciicast error}
rs <- r_session$new()
rs$run(function() library("not-a-package"))
```

callr also adds two stack traces to the output, one for the main process
and one for the subprocess:

```{asciicast last-error}
.Last.error
```

Errors from a `$call()` are returned in the `error` entry of the result:

```{asciicast error-call}
rs$call(function() library("still-not"))
rs$poll_process(2000)
rs$read()
```

## Debugging

Debugging subprocesses is hard.
`r_session` objects have a couple of methods to help you, but it is still
hard.

### Stack traces

As you have seen above, callr returns stack traces for errors, both for the
main process and the subprocess.
If your packages are installed with source references, then these include
links to the source files as well.

### `.Last.error`

For errors that are re-thrown in the main process, callr sets the
`.Last.error` variable to the last error object. You can inspect that after
the error.

`.Last.error$parent` contains the error object from the subprocess.
The error object often has additional information about the error, e.g.
`processx::run()` includes the standard output + error if the system process
exits with a non-successful status:

```{asciicast last-error-parent}
rs <- r_session$new()
rs$run(function() processx::run("ls", "/not-this"))
```

```{asciicast last-error-parent-1}
.Last.error
```

```{asciicast last-error-parent-2}
.Last.error$parent
```

### Inspecting the stack traces

Another way to inspect the stack trace in the subprocess is to set the
`callr.traceback` option to `TRUE` and call the `$traceback()` method after
the error.

This option is off by default, because the stack trace sometimes contains
large objects, that take a lot of time to copy between processes.


```{asciicast base-trace}
options(callr.traceback = TRUE)
rs <- r_session$new()
fun <- function() {
  options(warn = 2)         # convert warnings to errors
  f1 <- function() f2()
  f2 <- function() f3()
  f3 <- function() {
    vec <- 1:2
    if (vec) "success"
  }
  f1()
}
rs$run(fun)
```

```{asciicast base-trace-1}
rs$traceback()
```

### Inspecting frames of a stack trace

If the `callr.traceback` option is `TRUE`, then callr saves the full trace,
including the frames.
You can then inspect these frames with the `$debug()` method.
We can use it here to debug the previous error:

```{asciicast debug}
#| R.options = utils::modifyList(asciicast::asciicast_knitr_options(),
#|                               list(asciicast_at = NULL,
#|                                    asciicast_empty_wait = 5,
#|                                    asciicast_end_wait = 20,
#|                                    asciicast_omit_last_line = TRUE,
#|                                    asciicast_cols = 80,
#|                                    width = 80,
#|                                    asciicast_cursor = TRUE,
#|                                    asciicast_incomplete_error = FALSE)
#|                              ),
#| echo = FALSE,
#| asciicast_echo = TRUE
rs$debug()
#! --
.inspect 3
#! --
ls()
#! --
vec
#! --
.q
```

### Interactive debugging

You can use the `$attach` method to start a REPL (read-eval-print loop)
that runs in the subprocess.
It is best to do this when the subprocess is idle, otherwise it is
probably not responsive.

Press CTRL+C or ESC, or type `.q` and press ENTER to quit the REPL.

Here is an example:

```{asciicast attach}
rs <- r_session$new()
rs$run(function() { .GlobalEnv$data <- mtcars; NULL })
```

```{asciicast attach-1}
#| R.options = utils::modifyList(asciicast::asciicast_knitr_options(),
#|                               list(asciicast_at = NULL,
#|                                    asciicast_empty_wait = 5,
#|                                    asciicast_end_wait = 20,
#|                                    asciicast_omit_last_line = TRUE,
#|                                    asciicast_cols = 80,
#|                                    width = 80,
#|                                    asciicast_cursor = TRUE,
#|                                    asciicast_incomplete_error = FALSE)
#|                              ),
#| echo = FALSE,
#| asciicast_echo = TRUE
rs$attach()
#! --
ls()
#! --
head(data)
#! --
.q
```



This is an experimental feature and it does not always print the output
properly, e.g. sometimes you need to press ENTER twice, but nevertheless
it can be useful at times.

## Communication protocol

The `$read()` method can return messages with the following `code`s:

* `200`: the function is done. Note that the result might still be an
  error, you need to check that the `error` entry is not `NULL`.
* `201`: the R process is ready to use. This is the first message you
  get after a successful non-blocking startup.
* `202`: attach is done. This is used internally by the `$attach()` method,
  see the section about debugging below.
* `301`: message from the subprocess. E.g. the cli package can generate
  such messages, see
  [the cli documentation](https://cli.r-lib.org/articles/semantic-cli.html#sub-processes).
* `500`: the R session exited cleanly. This means that the evaluated
  expression quit R.
* `501`: the R session crashed or was killed.
* `502`: the R session closed its end of the connection that callr uses
  for communication. This might also happen because it was killed or
  crashed.
